\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

I would like to thank my supervisor, Rob Noble. Needless to say that none of the
results presented in this thesis would have been possible without his guidance.
More than that, however, I would like to thank him for his patience, as I came
to him with a possibly negative amount of knowledge about mathematical biology
in the middle of the worst pandemic in a century. I cannot imagine myself seeing
a project of this scale without his help, and could not have asked for a better
supervisor.\par
I would also like to thank my comrades at City maths, who let me drag them away
from their work and would listen to me complain about work, politics, life, and
everything in between. The daily poco poco at Cafe Fiori is a ritual I will
have a hard time getting used to not having. \par
Outside City, I am grateful to my flatmates, past and present, who have also
listened to hours of my rambling and have been a constant source of support and
inspiration through their own work. Alex especially has come in clutch with some
of the more obscure references in branching processes that let me bring this
thesis to a satisfying conclusion. \par
And of course, I am grateful to my parents for their daily messages and calls,
reminding me that I have the support of a loving family behind me. Finally, a
special thanks to Ana, the smartest person I've ever met, and the love of my
life, whose turn it is now to be the spoiled PhD student complaining how
difficult it is to write a thesis. \par
